### 2017-12-07 manuscript.rmd
incorporated Casey's changes

This manuscript is an executable document computed directly from the data, providing an explicit and reproducible description of all findings. All scripts for the analyses are available in a git repository at [https://github.com/caseywdunn/siphonophore_phylogeny_2017](https://github.com/caseywdunn/siphonophore_phylogeny_2017). The most recent commit at the time of the analysis presented here was `r system("git log | head -n 1", intern=TRUE) %>% str_replace("commit ", "")`.

@@

Collection data on all examined specimens, a description of the tissue that was sampled from the colony, collection mode, sample processing details, mRNA extraction methods, sequencing library preparation methods and sequencing details are summarized in supplementary table 1. Monterey Bay and Gulf of California specimens were collected by remotely operated underwater vehicle (ROV) or during blue-water scuba dives. *Chelophyes appendiculata* and *Hippopodius hippopus* specimens were collected in the bay of Villefranche-sur-Mer, France, during a plankton trawl on 04/13/11. Available physical vouchers have been deposited at the Museum of Comparative Zoology (Harvard University), Cambridge, MA, or had been previously deposited at the United States National Museum (Smithsonian Institution), Washington, DC. Accession numbers are given in supplementary table 1. In cases where physical vouchers were unavailable we provide photographs to document species identity (see git repository).

@@

When possible specimens were starved overnight in filtered seawater at temperatures close to ambient water temperatures at the time point of specimen collection (see supplementary document). mRNA was extracted directly from tissue using a variety of methods (see supplementary document): Magnetic mRNA Isolation Kit (NEB, #S1550S), Invitrogen Dynabeads mRNA Direct Kit (Ambion, #61011), Zymo Quick RNA MicroPrep (Zymo #R1050), or from total RNA after Trizol (Ambion, #15596026) extraction and through purification using Dynabeads mRNA Purification Kit (Ambion, #61006)-  in case of anticipated very small total RNA quantities, only a single round of bead purification was performed; or Trizol directly into the Illumina TruSeq Stranded Library Kit. Extractions were performed according to the manufacturer's instruction.  Any resulting higher rRNA read counts were dealt with further downstream in the bioinformatics workflow. Libraries were prepared for sequencing using the Illumina TruSeq RNA Sample Prep Kit (Illumina, #FC-122-1001, #FC-122-1002), the Illumina TruSeq Stranded Library Prep Kit (Illumina, #RS-122-2101) or the NEBNext RNA Sample Prep Master Mix Set (NEB, #E6110S). We collected long read paired end Illumina data for *de novo* transcriptome assembly. In the case of large tissue inputs, libraries were sequenced separately for each tissue, subsequently subsampled and pooled *in silico*. Libraries were sequenced on the HiSeq 2000, 2500, and 3000 sequencing platforms (see supplementary document).

@@

New data were analysed in conjunction with 13 publically available datasets, with a total number of 43 species. Sequence assembly, annotation, Maximum Likelihood (ML) phylogenetic analysis were conducted with the tool Agalma v. 1.00 [@Dunn:2013kw; @Guang202416], and Bayesian Inference (BI) analyses were conducted using Phylobayes v. 1.7a-mpi [@lartillot2009phylobayes]. Source code for all analysis steps, sequence alignments, sampled and consensus trees, and voucher information are available in a git repository https://github.com/caseywdunn/siphonophore_phylogeny_2017.

Two outgroup species, *Atolla vanhoeffeni* and *Aegina citrea*, were removed from the final supermatrix and phylogeny due to low gene occupancy (gene sampling of 20.8% and 14.5% respectively in a 50% occupancy matrix with 2,203 genes). ML analyses were conducted on the unpartitioned supermatrix using the WAG+$\Gamma$ model of amino acid substitution, and bootstrap values were estimated using 1000 replicates. BI was conducted using two different CAT models, CAT-Poisson and CAT-GTR [@doi:10.1093/molbev/msh112]. Two independent MCMC chains were run under the CAT-GTR model, and four independent MCMC chains were run under the CAT-Poisson model. The CAT-GTR and CAT-Poisson models did not converge after a long CPU time, and only the results from the CAT-Poisson model are included here.

Morphological character data used in trait mapping were obtained from the literature, or from direct observation of available voucher material. Depth distribution data was queried from the MBARI VARS ( http://www.mbari.org/products/research-software/video-annotation-and-reference-system-vars/, [@schlining2006mbari]) database (with assistance and permission from SHD Haddock), including records from several MBARI cruises. We used stochastic character mapping to infer the probable evolution of traits on the tree in R using the `phytools` package [@MEE3:MEE3169; @huelsenbeck2003stochastic]. Subsequent analyses were conducted in R and integrated into this manuscript with the `knitr` package. See Supplementary Information for R package version numbers.

#### Hypothesis testing
We used the Swofford-Olsen-Waddell-Hillis (SOWH) test [@swofford1996phylogenetic] to evaluate two hypotheses: (i) physonects are monophyletic [@totton1965synopsis]; (ii) monoecious species are monophyletic [@Dunn:2005dy]. As the sexual system of *Rudjakovia* sp. is unclear, we carried out two tests of the monophyly of monoecy, one with *Rudjakovia* sp. included as a monoecious species, and one without. We used SOWHAT [@doi:10.1093/sysbio/syv055] dev. version 0.39 (commit fd68ef5733c095c7000a4f92dc8c0daaddeec3b9) to carry out the SOWH tests in parallel with the default options and an initial sample size of 100 (source code can be found in the git repository). For each hypothesis we defined a topology with a single constrained node that was inconsistent with the most likely topology (Fig. 3 x). We used a threshold for significance of 0.05 and following the initial 100 samples, we evaluated the confidence interval around the p-value to determine if more samples were necessary.

@@

### Species phylogeny and hypothesis testing
Specimens were collected in the eastern Pacific Ocean, Mediterranean, and the Gulf of California (table 1). All sequence data have been deposited in the sequence read archive (SRA). The analyses presented here consider 33 siphonophore species and 8 outgroup species. This includes new data for 30 species. Summary statistics for expression libraries are given in supplementary table x. In the final analyses, we sampled 1,071 genes to generate a supermatrix with 60% occupancy and a length of 378,468 amino acids (gene occupancy matrix - supplementary figure 1). 
```{r specimen_summary, results='asis', echo=FALSE, comment=NA, warning=FALSE}
specimen_data <- read.csv("supplementary_info/Specimen_data.csv")
names(specimen_data)<-c("New data","Species","ID","SRA Number", "Library ID", "Library type","Tissue","Sequencer", "Sequencing center", "Sample prep","Collected by","Collection date","Specimen id","Depth (m)","Geo loc name","Lat Lon","Voucher Type","Voucher location","Voucher Cat No.","Collection mode")

@@

```
Maximimum likelihood analyses had `r length(bootstrap_trees)` replicates. We ran `r length( input_files_phylobayes )` phylobayes chains, and visual inspection of the traces indicated that a burn in of `r burn_in` trees was sufficient for all runs. This left `r length( phylobayes_trees )` trees in the posterior. The phylobayes chains did not converge after a long CPU time, and uncertainty remains around the placement of *Erenna richardi*, and also *Nanomia bijuga*. Within the ML analysis, the placement of *Erenna richardi* is also unstable. Alternative topologies for these nodes are shown in figure 3 x.

@@

```
These findings are largely consistent with a previous analysis based on two genes (16S and 18S ribosomal RNA) (Fig. 3) [@Dunn:2005dy]. There is strong support for the Cystonectae as sister to all other siphonophores, the Codonophora. We also find strong support for Calycophorae nested within the paraphyletic "Physonectae", and for the Apolemidae as sister to all other codonophorans. Multiple nodes that were not resolved in the previous two-gene analysis receive strong support in this 1,071-gene transcriptome analysis. There is strong support for the Pyrostephidae as sister to all other non-Apolemid codonophorans. Within the clade that is sister to the Pyrostephidae, we find two main clades, the Calycophorae and a clade we call the Euphysonectae, which includes the remaining non-Apolemid, non-Pyrostephid "Physonectae". Given the relatively shallow sampling of this analysis, we define the Euphysonectae as the clade consisting of *Agalma elegans* and all taxa that are more closely related to it than to the Calycophorae.

In this phylogeny, *Physophora gilmeri* along with *Lychnagalma utricularia* (both not included in the previous phylogeny) are sister to the Agalmatidae *sensu stricto*, a clade restricted to *Agalma*, the Athorybiidae, *Halistemma* and *Nanomia* [@Dunn:2005dy; @pugh2006taxonomic]. In the rDNA study, *Physophora hydrostatica* (sister species to *P. gilmeri*) was sister to the Forskaliidae with low support. *Cordagalma cordiforme* (= *Cordagalma ordinatum*) [@pugh2016synopsis] was previously unresolved, while in this analysis *Cordagalma* sp. is in a clade with *Forskalia asymmetrica*, falling outside of the Agalmatidae *sensu stricto*. Placement of *Cordagalma* outside Agalmatidae *sensu stricto* is consistent with previous analyses using morphological and molecular data [@pugh2006taxonomic;@Dunn:2005dy].

Within the calycophorans, taxon sampling is shallower, however there is broad agreement with the previous analysis. Calycophorans have in the past been split into convenient groupings of prayomorph and diphyomorph based on morphology (after @mackie1987siphonophore). As in the previous study, this phylogeny indicates that the prayomorphs are paraphyletic. *Craseoa lathetica* and *Desmophyes* sp. are sister to *Hippopodius hippopus* in this study, while in the previous study, the relationship between *Craseoa lathetica* and the clade including *Hippopodius hippopus* was unresolved.

@@

## Character Evolution

The morphological character matrix, and all the code used to generate the character mapping is available in the git repository and in the supplementary materials. A  number of traits were mapped to tips on the phylogeny, including: the presence/absence of palpons, tentilla, bracts, pneumatophores, nectosome, and eudoxia, monoecy vs dioecy, the orientation of nectosome relative to siphosome, as well as the median vertical distribution (Fig. 4, 5, S2-5).

```{r char_evo_setup, include=F, echo=F}
#Load data
read.csv('character_coding/main_characters.csv', header = T, sep = ',') -> cdata
rownames(cdata) = cdata$Species

@@

#Descending mantle Canal SIMMAP Setup
desc_mantle_data = cdata[which(!is.na(cdata$Desc.Mantle.Canal)),]
desc_mantle = desc_mantle_data$Desc.Mantle.Canal
names(desc_mantle)=rownames(desc_mantle_data)
DescMantleTree =  drop.tip(ultratree, which(!(ultratree$tip.label %in% names(desc_mantle))))
make.simmap(DescMantleTree, desc_mantle, nsim = 100, message=FALSE) -> desc_mantle_sim

#Ascending Mantle canal SIMMAP Setup
asc_mantle_data = cdata[which(!is.na(cdata$Asc.Mantle.Canal)),]
asc_mantle = asc_mantle_data$Asc.Mantle.Canal
names(asc_mantle)=rownames(asc_mantle_data)
AscMantleTree =  drop.tip(ultratree, which(!(ultratree$tip.label %in% names(asc_mantle))))
make.simmap(AscMantleTree, asc_mantle, nsim = 100, message=FALSE) -> asc_mantle_sim
```

@@

#Simmap SEX
plotTree(ultratree, lwd = 4)
colors = c("black", "red")
names(colors) = c("Dioecious", "Monoecious")
sex_sim %>% plotSimmap(lwd = 4, add = T, colors=colors)
nodelabels(pie=(describe.simmap(sex_sim, plot=F)$ace) ,piecol=colors,cex=0.35)
add.simmap.legend(colors = colors, x=0.6*par()$usr[1],y=0.3*par()$usr[4],prompt=FALSE)
sex_obj <- densityMap(sex_sim)
sex_obj$cols[1:length(sex_obj$cols)]<-colorRampPalette(colors=c("blue", "orange"), space="Lab")(length(sex_obj$cols))
 pdf("figures_raw/figure_5_raw_sex.pdf")
 plot(sex_obj)
 dev.off()
```

```{r palpons, echo=F, message=F, warning=F, include=F}
#Simmap PALPONS
 plotTree(ultratree, lwd = 4)
 colors = c("black", "red")
 names(colors) = c("Absent", "Present")
 palpons_sim %>% plotSimmap(lwd = 4, add = T,colors=colors)
 nodelabels(pie=(describe.simmap(palpons_sim, plot=F)$ace) ,piecol=colors,cex=0.35)
 add.simmap.legend(colors = colors, x=0.6*par()$usr[1],y=0.3*par()$usr[4],prompt=FALSE)
palpons_obj <- densityMap(palpons_sim)
palpons_obj$cols[1:length(palpons_obj$cols)]<-colorRampPalette(colors=c("blue", "orange"), space="Lab")(length(palpons_obj$cols))
 pdf("figures_raw/figure_5_raw_palpons.pdf")
 plot(palpons_obj)
 dev.off()
```

@@

```{r ggtree_bin_heatmap, echo=F, message=F, warning=F, include=F}
p <- ggtree(ultratree)
p <- p + geom_tiplab(size=2.2)
f_cdata = sapply(cdata, as.factor) %>% as.data.frame %>% .[c(1,2,5:9,17,18,3)]
rownames(f_cdata) = rownames(cdata)
f_cdata = f_cdata[,c(7,4,1,3,6,2,10,8,9,5)]
f_cdata[match(tree$tip.label,rownames(f_cdata)),]

@@

Siphonophore colonies are either monoecious (male and female gonophores are on the same colony) or dioecious (male and female gonophores are on different colonies). Previous analyses suggested that the common ancestor of siphonophores was dioecious, and there is a single gain of monoecy within the Codonophora; however the lack of resolution of deep relationships within Codonophora made it impossible to rule out alternative evolutionary scenarios [@Dunn:2005dy]. The current better-resolved tree (Fig. 5A), indicates that monoecy in siphonophores evolved twice from a dioecious ancestor, in the branch leading to Calycophorae and in the branch leading to [**placeholder name** Agalmatids *sensu lato*]. There is a small probability for an alternative scenario featuring a single gain of monoecy before the split of the Codonophora, with a subsequent shift to dioecy in the *Marrus-Erenna* clade.

@@

One of the most striking aspects of siphonophore biology is their diversity of unique zooid types [@Beklemishev1969;@cartwright2010character]. For example, the genus *Forskalia* has 6 basic zooid types (pneumatophore, nectophore, gastrozooid, palpon, bract, and gonophore), and a total of 10 counting subtypes (4 types of bract, male & female gonophores)[@pugh2003revision]. Diphyomorphs have more than 1 type of propulsive nectophore, while cystonects have none. Here we reconstruct the evolutionary origins of the different zooid types on the present transcriptome tree (Fig. 4).

Nectophores are retained modified medusae that codonophorans use for coordinated colony-level swimming. The nectosome is the region of the colony that develops from the nectosomal growth zone (Fig. 2B). Unlike the siphosomal growth zone, the nectosomal growth zone does not bud gastrozooids, but nectophores (and in the case of *Apolemia*, also palpons). It is possible that the common ancestor of siphonophores had a nectosome, which was lost on the branch leading to cystonects. We cannot exclude with certainty the alternative hypothesis of a nectosome-less ancestor followed by a gain of the nectosome in the branch leading to Codonophora. It is suggested that the nectosome arose as a duplication of the siphosome, followed by functional specialization in propelling the colony [@dunn2006evolution]. The nectosome has been lost within Codonophora in the genus *Athorybia*.

@@

Following the colony development orientation framework [@haddock_dunn_pugh_2005], the nectosome can be located in a dorsal or a ventral position. Our ancestral reconstructions for this character (Fig. S5) show that a ventrally-oriented nectosome was the ancestral form in siphonophores, and that a dorsal nectosome has evolved twice independently, in the branch leading to the Agalmatidae *sensu stricto* and the branch leading to the *Bargmannia* species.

Bracts are highly reduced zooids unique to siphonophores, but they are only present in Codonophora (Fig. 4). Bracts are functional for protection of the delicate zooids and to help maintain neutral buoyancy [@jacobs1937beobachtungen]. Some calycophorans are able to actively exclude sulphate ions in their bracts to adjust their buoyancy along the colony [@bidigare1980role]. As with the nectosome, we cannot be sure whether the common ancestor of siphonophores had bracts or not. The common ancestor of Codonophora had only one bract subtype, which was lost in Hippopodidae, *Physophora hydrostatica*, the Clausophyidae, and in *Gymnopraia lapislazula* (none of the latter three species and clades are included in the phylogeny). Phylogenetic reconstructions are sensitive to taxon sampling, and these patterns of loss are not captured by the present sampling [**We don't have the reconstruction in supp info - do we want to add it?**].

Palpons are modified reduced gastrozooids used for digestion and circulation of the gastrovascular fluid [@mackie1987siphonophore]. We do not distinguish here between gonopalpons (palpons associated with gonodendra, without a tentacle, as in the cystonects) and palpons borne on the stem (typically with a reduced tentacle or palpacle) [@totton1965synopsis]. They were present in the common ancestor of siphonophores (Fig. 5B), retained in most species, but lost three times independently in the branches leading to *Pyrostephidae* (represented here by the genera *Bargmannia* and *Rudjakovia*), in calycophorans (with the exception of Stephanophyes [@totton1965synopsis]), and in *Marrus claudanielis*. These taxa might have found other avenues to effectively circulate nutrients across the colony.

@@

The pneumatophore is a gas filled float located at the anterior end of the colony, which helps the colony maintain its orientation in the water column, and plays a role in flotation in the case of the cystonects [@church2015histology;@mackie1974locomotion; @totton1965synopsis]. Recent evidence of neural arrangement in the pneumatophore of *Nanomia bijuga*  suggests it could also gather information on relative pressure changes (and thus depth changes), helping regulate geotaxis [@church2015histology]. The ancestral siphonophore had a pneumatophore (Fig. 2B), since both cystonects and most codonophorans possess one (Fig. 4). It is not considered a zooid, as it is not formed by budding but by invagination at the aboral end of the planula during early development [@garstang1946morphology; @leloup1935siphonophores; @carre1969etude].  The pneumatophore was lost in Calycophorae and never gained again in that clade. Calycophorans rely on the ionic balance of their gelatinous nectophores and bracts to retain posture and neutral buoyancy [@mackie1974locomotion].

@@

The most complex nematocyst batteries of Cnidaria can arguably be found among the siphonophores, hanging in regularly spaced tentacle side branches called tentilla (Fig. 2A). Most hydrozoans, including the clade that contains siphonophores, bear simple tentacles (tentacles with no side branches). It is still an open question whether the common ancestor of Siphonophora had simple or branched tentacles. The only siphonophores species regarded as lacking tentilla are *Physalia physalis*, *Apolemia* spp., and *Bathyphysa conifera*. Since *B. conifera* is the only member of the *Rhizophysidae* (and of the *Bathyphysa* genus) lacking tentilla, we can safely assume this is a case of secondary loss. When we reconstruct the evolution of this character on the current phylogeny, 70% of simulations support an common ancestor bearing tentilla, with two independent losses leading to *Physalia* and *Apolemia* (Fig. S2). However, this leaves a 30% support for a simple-tentacled common ancestor followed by 2 independent gains of tentilla in the branches leading to *Rhizophysidae* and non-apolemiid codonophorans.

>>>>

All scripts for the analyses are available in a git repository at [https://github.com/caseywdunn/siphonophore_phylogeny_2017](https://github.com/caseywdunn/siphonophore_phylogeny_2017). The most recent commit at the time of the analysis presented here was `r system("git log | head -n 1", intern=TRUE) %>% str_replace("commit ", "")`.

@@

Specimens were collected in the eastern Pacific Ocean, Mediterranean, and the Gulf of California.Collection data on all examined specimens, a description of the tissue that was sampled from the colony, collection mode, sample processing details, mRNA extraction methods, sequencing library preparation methods and sequencing details are summarized in Supplementary Information. Monterey Bay and Gulf of California specimens were collected by remotely operated underwater vehicle (ROV) or during blue-water SCUBA dives. *Chelophyes appendiculata* and *Hippopodius hippopus* specimens were collected in the bay of Villefranche-sur-Mer, France, during a plankton trawl on 13 April 2011. Available physical vouchers have been deposited at the Museum of Comparative Zoology (Harvard University), Cambridge, MA, or had been previously deposited at the United States National Museum (Smithsonian Institution), Washington, DC. Accession numbers are given in Supplementary Information. In cases where physical vouchers were unavailable we provide photographs to document species identity (see git repository).

@@

When possible, specimens were starved overnight in filtered seawater at temperatures close to ambient water temperatures at the time point of specimen collection (see supplementary document). mRNA was extracted directly from tissue using a variety of methods (see supplementary document): Magnetic mRNA Isolation Kit (NEB, #S1550S), Invitrogen Dynabeads mRNA Direct Kit (Ambion, #61011), Zymo Quick RNA MicroPrep (Zymo #R1050), or from total RNA after Trizol (Ambion, #15596026) extraction and through purification using Dynabeads mRNA Purification Kit (Ambion, #61006).  In case of anticipated very small total RNA quantities, only a single round of bead purification was performed; or Trizol extraction of total RNA directly into the Illumina TruSeq Stranded Library Kit. Extractions were performed according to the manufacturer's instruction. All samples were DNase treated (TURBO DNA-free, Invitrogen #AM1907;or on column DNase treatment with Zymo Quick RNA MicroPrep). Libraries were prepared for sequencing using the Illumina TruSeq RNA Sample Prep Kit (Illumina, #FC-122-1001, #FC-122-1002), the Illumina TruSeq Stranded Library Prep Kit (Illumina, #RS-122-2101) or the NEBNext RNA Sample Prep Master Mix Set (NEB, #E6110S). We collected long read paired end Illumina data for *de novo* transcriptome assembly. In the case of large tissue inputs, libraries were sequenced separately for each tissue, subsequently subsampled and pooled *in silico*. Libraries were sequenced on the HiSeq 2000, 2500, and 3000 sequencing platforms.

@@

New data were analysed in conjunction with 13 publically available datasets, with a total number of 43 species. Sequence assembly, annotation, and Maximum Likelihood (ML) phylogenetic analysis were conducted with the tool Agalma v. 1.00 [@Dunn:2013kw; @Guang202416], and Bayesian Inference (BI) analyses were conducted using Phylobayes v. 1.7a-mpi [@lartillot2009phylobayes]. Sequence alignments, sampled and consensus trees, and voucher information are available in a git repository https://github.com/caseywdunn/siphonophore_phylogeny_2017. All sequence data have been deposited in the NCBI sequence read archive (SRA).

Two outgroup species, *Atolla vanhoeffeni* and *Aegina citrea*, were removed from the final supermatrix due to low gene occupancy (gene sampling of 20.8% and 14.5% respectively in a 50% occupancy matrix with 2,203 genes). ML analyses were conducted on the unpartitioned supermatrix using the WAG+$\Gamma$ model of amino acid substitution, and bootstrap values were estimated using 1000 replicates. BI was conducted using two different CAT models, CAT-Poisson and CAT-GTR [@doi:10.1093/molbev/msh112]. Two independent MCMC chains were run under the CAT-GTR model, and four independent MCMC chains were run under the CAT-Poisson model. The CAT-GTR and CAT-Poisson models did not converge, and only the results from the CAT-Poisson model are included here. The analyses presented here consider 33 siphonophore species and 8 outgroup species. This includes new data for 30 species. Summary statistics for expression libraries are given in Supplementary Information. In the final analyses, we sampled 1,071 genes to generate a supermatrix with 60% occupancy and a length of 378,468 amino acids (gene occupancy matrix - supplementary figure 1).

Maximimum likelihood analyses had `r length(bootstrap_trees)` replicates. We ran `r length( input_files_phylobayes )` phylobayes chains, and visual inspection of the traces indicated that a burn in of `r burn_in` trees was sufficient for all runs. This left `r length( phylobayes_trees )` trees in the posterior. The phylobayes chains did not converge, and uncertainty remains around the placement of *Erenna richardi*, and also *Nanomia bijuga*. Within the ML analysis, the placement of *Erenna richardi* is also unstable. Alternative topologies for these nodes are shown in figure 3.

Morphological character data used in trait mapping were obtained from the literature or direct observation of available voucher material. Depth distribution data was queried from the MBARI VARS ( http://www.mbari.org/products/research-software/video-annotation-and-reference-system-vars/, [@schlining2006mbari]) database. We used stochastic character mapping to infer the probable evolution of traits on the tree in R using the `phytools` package [@MEE3:MEE3169; @huelsenbeck2003stochastic]. Subsequent analyses were conducted in R and integrated into this manuscript with the `knitr` package. See Supplementary Information for R package version numbers.

#### Hypothesis testing
We used the Swofford-Olsen-Waddell-Hillis (SOWH) test [@swofford1996phylogenetic] to evaluate two hypotheses: (i) physonects are monophyletic [@totton1965synopsis]; (ii) monoecious species are monophyletic [@Dunn:2005dy]. As the sexual system of *Rudjakovia* sp. is unclear, we carried out two tests of the monophyly of monoecy, one with *Rudjakovia* sp. included as a monoecious species, and one without. We used SOWHAT [@doi:10.1093/sysbio/syv055] dev. version 0.39 (commit fd68ef57) to carry out the SOWH tests in parallel with the default options and an initial sample size of 100 (source code can be found in the git repository). For each hypothesis we defined a topology with a single constrained node that was inconsistent with the most likely topology (Fig. 3). We used a threshold for significance of 0.05 and following the initial 100 samples, we evaluated the confidence interval around the p-value to determine if more samples were necessary.

@@

### Species phylogeny and hypothesis testing

```{r specimen_summary, results='asis', echo=FALSE, comment=NA, warning=FALSE}
specimen_data <- read.csv("supplementary_info/Specimen_data.csv")
names(specimen_data)<-c("New data","Species","ID","SRA Number", "Library ID", "Library type","Tissue","Sequencer", "Sequencing center", "Sample prep","Collected by","Collection date","Specimen id","Depth (m)","Geo loc name","Lat Lon","Voucher Type","Voucher location","Voucher Cat No.","Collection mode")

@@

@@

```
These findings are largely consistent with a previous analysis based on two genes (16S and 18S ribosomal RNA) (Fig. 3) [@Dunn:2005dy]. There is strong support for the Cystonectae as sister to all other siphonophores, the Codonophora. We also find strong support for Calycophorae nested within the paraphyletic "Physonectae", and for the Apolemidae as sister to all other codonophorans. Multiple nodes that were not resolved in the previous two-gene analysis receive strong support in this 1,071-gene transcriptome analysis. There is strong support for the Pyrostephidae as sister to all other non-Apolemid codonophorans. Within the clade that is sister to the Pyrostephidae, we find two main clades, the Calycophorae and a clade we call the Euphysonectae (Fig. 3), which includes the remaining non-Apolemid, non-Pyrostephid "Physonectae". We define the Euphysonectae as the clade consisting of *Agalma elegans* and all taxa that are more closely related to it than to *Diphyes dispar*. Within Euphysonectae, we identify Clade A and Clade B (Fig. 3). Clade A is a group of siphonophores that all have an involucrum, or fold, that forms around the base of the tentilla [@totton1965synopsis]. Clade B is a group of deep-sea siphonophores found within the bathypelagic. Within Clade B there is low support for the placement of *Erenna richardi*, and more taxon sampling will be required to determine the relationship of species within this clade.

Within Clade A, *Physophora gilmeri* along with *Lychnagalma utricularia* (both not included in the previous phylogeny) are sister to the Agalmatidae *sensu stricto*, a clade restricted to *Agalma*, the Athorybiidae, *Halistemma* and *Nanomia* [@Dunn:2005dy; @pugh2006taxonomic]. In the rDNA study, *Physophora hydrostatica* (sister species to *P. gilmeri*) was sister to the Forskaliidae with low support. The position of *Cordagalma cordiforme* (= *Cordagalma ordinatum*) [@pugh2016synopsis] was previously unresolved, while in this analysis *Cordagalma* sp. is in a clade with *Forskalia asymmetrica*, falling outside of the Agalmatidae *sensu stricto*. Placement of *Cordagalma* outside Agalmatidae *sensu stricto* is consistent with previous analyses using morphological and molecular data [@pugh2006taxonomic;@Dunn:2005dy].

Within the calycophorans, taxon sampling is shallower, however there is broad agreement with the previous analysis. Calycophorans have in the past been split into convenient groupings of prayomorph and diphyomorph based on morphology (after @mackie1987siphonophore). As in the previous study, this phylogeny indicates that the prayomorphs are paraphyletic with respect to the diphyomorphs. *Craseoa lathetica* and *Desmophyes* sp. are sister to *Hippopodius hippopus* in this study, while in the previous study, the relationship between *Craseoa lathetica* and the clade including *Hippopodius hippopus* was unresolved.

@@

## Character Evolution
```{r char_evo_setup, include=F, echo=F}
#Load data
read.csv('character_coding/main_characters.csv', header = T, sep = ',') -> cdata
rownames(cdata) = cdata$Species

@@

#Mantle canal SIMMAP Setup
#mantle_data = cdata[which(!is.na(cdata$Mantle.Canal)),]
#mantle = mantle_data$Mantle.Canal
#names(mantle)=rownames(mantle_data)
#MantleTree =  drop.tip(ultratree, which(!(ultratree$tip.label %in% names(mantle))))
#make.simmap(MantleTree, mantle, nsim = 100, message=FALSE) -> mantle_sim
```

@@

#Simmap SEX
plotTree(ultratree, lwd = 4)
colors = c("black", "red")
names(colors) = c("Dioecious", "Monoecious")
sex_sim %>% plotSimmap(lwd = 4, add = T, colors=colors)
nodelabels(pie=(describe.simmap(sex_sim, plot=F)$ace) ,piecol=colors,cex=0.35)
add.simmap.legend(colors = colors, x=0.6*par()$usr[1],y=0.3*par()$usr[4],prompt=FALSE)
 pdf("figures_raw/figure_5_raw_sex.pdf")
 densityMap(sex_sim)
 dev.off()
```

```{r palpons, echo=F, message=F, warning=F, include=F}
#Simmap PALPONS
 plotTree(ultratree, lwd = 4)
 colors = c("black", "red")
 names(colors) = c("Absent", "Present")
 palpons_sim %>% plotSimmap(lwd = 4, add = T,colors=colors)
 nodelabels(pie=(describe.simmap(palpons_sim, plot=F)$ace) ,piecol=colors,cex=0.35)
 add.simmap.legend(colors = colors, x=0.6*par()$usr[1],y=0.3*par()$usr[4],prompt=FALSE)
 pdf("figures_raw/figure_5_raw_palpons.pdf")
 densityMap(palpons_sim)
 dev.off()
```

@@

```{r ggtree_bin_heatmap, echo=F, message=F, warning=F, include=F}
p <- ggtree(ultratree)
p <- p + geom_tiplab(size=2.2)
f_cdata = sapply(cdata, as.factor) %>% as.data.frame %>% .[c(1,2,5:9)]
rownames(f_cdata) = rownames(cdata)
f_cdata = f_cdata[,c(4,7,1,3,6,2,5)]
f_cdata[match(tree$tip.label,rownames(f_cdata)),]

@@

Siphonophore colonies are either monoecious (male and female gonophores are on the same colony) or dioecious (male and female gonophores are on different colonies). The current better-resolved tree (Fig. 5A), indicates that monoecy in siphonophores evolved twice from a dioecious ancestor, in the branch leading to Calycophorae and in the branch leading to Clade A. There is a small probability for an alternative scenario featuring a single gain of monoecy before the split of the Codonophora, with a subsequent shift to dioecy in the *Marrus-Erenna* clade. Previous analyses suggested that the common ancestor of siphonophores was dioecious, and there is a single gain of monoecy within the Codonophora [@Dunn:2005dy].

@@

One of the most striking aspects of siphonophore biology is their diversity of unique zooid types [@Beklemishev1969;@cartwright2010character]. For example, the genus *Forskalia* has 6 basic zooid types (pneumatophore, nectophore, gastrozooid, palpon, bract, and gonophore), and a total of 10 when considering zooid subtypes (4 types of bract, male & female gonophores)[@pugh2003revision]. Diphyomorphs have more than one type of propulsive nectophore, while cystonects have none. Here we reconstruct the evolutionary origins of several zooid types on the present transcriptome tree (Fig. 4).

Nectophores are non-reproductive propulsive medusae that codonophorans use for coordinated colony-level swimming (Fig. 2). Nectophores are also present within the gonodendra (reproductive structures) of the cystonects, and are thought to propel the gonodendra when they detach from the colony [@totton1965synopsis;@totton1960]. It is not clear whether the nectophores found within the siphosome of the cystonects is homologous to the nectophores borne on the nectosome of the codonophora. The nectosome is the region of the colony that develops from the nectosomal growth zone (Fig. 2B). Unlike the siphosomal growth zone, the nectosomal growth zone does not bud gastrozooids, but nectophores (and in the case of *Apolemia*, also palpons). The present tree, as well as the analyses of @Dunn:2005dy, are consistent with a single origin of the nectosome.

@@

The nectosome can be located in a dorsal or a ventral position, following the orientation framework of @haddock_dunn_pugh_2005. Our ancestral reconstructions for this character (Fig. S5) show that a ventrally-oriented nectosome was the ancestral form in Codonophora, and that a dorsal nectosome has evolved twice independently, in the branch leading to the Agalmatidae *sensu stricto* and the branch leading to the *Bargmannia* species.

Bracts are highly reduced zooids unique to siphonophores, where they are only present in Codonophora (Fig. 4). Bracts are functional for protection of the delicate zooids and to help maintain neutral buoyancy [@jacobs1937beobachtungen]. Some calycophorans are able to actively exclude sulphate ions in their bracts to adjust their buoyancy along the colony [@bidigare1980role]. The common ancestor of Codonophora had only one bract subtype, which was lost in Hippopodidae, *Physophora hydrostatica*, the Clausophyidae, and in *Gymnopraia lapislazula* (none of the latter three species and clades are included in the phylogeny). Phylogenetic reconstructions are sensitive to taxon sampling, and these patterns of loss are not captured by the present sampling [**We don't have the reconstruction in supp info - do we want to add it?**].

Palpons are modified reduced gastrozooids used for digestion and circulation of the gastrovascular fluid [@mackie1987siphonophore]. We do not distinguish here between gonopalpons (palpons associated with gonodendra, without a tentacle, as in the cystonects) and palpons borne on the stem (typically with a reduced tentacle or palpacle) [@totton1965synopsis]. We reconstruct them as present in the common ancestor of siphonophores (Fig. 5B), retained in most species, but lost three times independently in the branches leading to *Pyrostephidae* (represented here by the genera *Bargmannia* and *Rudjakovia*), in calycophorans (with the exception of Stephanophyes [@totton1965synopsis]), and in *Marrus claudanielis*.

@@

The pneumatophore is a gas filled float located at the anterior end of the colony, which helps the colony maintain its orientation in the water column, and plays a role in flotation in the case of the cystonects [@church2015histology;@mackie1974locomotion; @totton1965synopsis]. It is not considered a zooid, as it is not formed by budding but by invagination at the aboral end of the planula during early development [@garstang1946morphology; @leloup1935siphonophores; @carre1969etude]. Recent evidence of neural arrangement in the pneumatophore of *Nanomia bijuga*  suggests it could also gather information on relative pressure changes (and thus depth changes), helping regulate geotaxis [@church2015histology]. The ancestral siphonophore had a pneumatophore (Fig. 2B), since both cystonects and most codonophorans possess one (Fig. 4).   The pneumatophore was lost in Calycophorae and never gained again in that clade. Calycophorans rely on the ionic balance of their gelatinous nectophores and bracts to retain posture and neutral buoyancy [@mackie1974locomotion].

@@

The most complex nematocyst batteries of Cnidaria can arguably be found among the siphonophores, hanging in regularly spaced tentacle side branches called tentilla (Fig. 2A). Most hydrozoans bear simple tentacles (tentacles with no side branches). It is still an open question whether the common ancestor of Siphonophora had tentilla. The only siphonophores species regarded as lacking tentilla are *Physalia physalis*, *Apolemia* spp., and *Bathyphysa conifera*. Since *B. conifera* is the only member of the *Rhizophysidae* (and of the *Bathyphysa* genus) lacking tentilla, we can safely assume this is a case of secondary loss. When we reconstruct the evolution of this character on the current phylogeny, 70% of simulations support an common ancestor bearing tentilla, with two independent losses leading to *Physalia* and *Apolemia* (Fig. S2). However, this leaves a 30% support for a simple-tentacled common ancestor followed by 2 independent gains of tentilla in the branches leading to *Rhizophysidae* and non-apolemiid codonophorans.
